---
title: "Initial_BCG_data_QC"
author: "Christy Dolph"
date: "2025-04-25"
output: html_document
editor_options: 
  chunk_output_type: console
---

# Overview: Preliminary Data Quality Checks

This script checks for obvious QC issues in the BCG dataset and generates a finalized data set for use in further analysis.

QC Issues include:

-   NA values

-   Creating additional identifiers

-   Adding flags for sites sampled by multiple agencies and/or multiple times in the same time period (year, month & year, exact same date)

-   Omit samples collected in winter (December, January, February) which are outside of typical range during which biotic condition is assessed

-   Calculate average BCG scores for sites sampled multiple times in one year, whether they were sampled either by multiple agencies or the same agency. Note that some of this duplication is because some state datasets were also included in the regional Chesapeake dataset

# 1 Set Up Workspace

Clean up workspace and load packages (or install packages if necessary)

```{r, setup, include =FALSE}
knitr::opts_chunk$set(echo = TRUE)

#Remove everything from environment:
rm(list=ls())

#Load required packages
want <- c("data.table", "fixest", "tidyverse", "tidylog", "readxl", "lubridate", "rmarkdown", "writexl", "sf", "mapview", "xtable", "pals", "tmap", "maps", "gridExtra", "ffmpeg", "magick", "animation", "geosphere", "nhdplusTools", "knitr")
   

need <- want[!(want %in% installed.packages()[,"Package"])]
if (length(need)) install.packages(need)
lapply(want, function(i) require(i, character.only=TRUE))
rm(want, need)
```

Set Up Working directories:

```{r, Create list of directories}
dir <- list()

#Main directory: 
dir$root <- getwd()

#Directory for loading data from EPA Project Round 1:
dir$data_r1 <- paste(dir$root, "/EPA_Round_1_data", sep = "")
#Directory for loading data from EPA Project Round 2:
dir$data_r2 <- paste(dir$root, "/EPA_Round_2_data", sep = "")
#Directory for spatial data: 
spatial_dir<-paste(dir$root, "/Spatial_data", sep="")
#Directory with HUC shapefiles: 
dir$data_huc  <- paste(dir$root, "/Spatial_data/HUC_spatial_data", sep = "")
#Output directory for most data: 
dir$output<-paste(dir$root, "/BCG_output", sep="")
```

# 2 Load Data Files

```{r}
data <- read_csv(paste(dir$output, "/EPA_BCG_allsamples_ALLROUNDS_NHDCOMID_clean_HUC12.csv", sep = ""))
setDT(data)

names(data)
nrow(data)
```

# 3 Generate Clean Tables for Analysis

-   Check for NA values in BCG or Lat/Long, remove if needed

-   Add year, month and day identifiers

-   Omit samples collected in winter (December, January, February) (i.e., outside of typical range during which biotic condition is assessed)

-   Add HUC2 identifiers

-   Add unique Site ID based on lat/long

-   Averages BCG across subsamples (multiple samples collected per site/day) and across samples per year, whether those samples were collected by the same or different agencies

    -   Note that some of this duplication is because some state datasets were also included in the regional Chesapeake or national EPA dataset

    -   Also calculate min and max BCG score in a given year

```{r}
#Check for NA values
colSums(is.na(data))

#remove observations with NA values for Lat & Long
nrow(data)
data <- data[!is.na(data$Lat) & !is.na(data$Long),]
nrow(data) #no rows removed

#rename columns to match script below
data<-data %>% 
  rename(COMID=comid)

names(data)

#Drop NAs for COMID
nrow(data)
data<-data %>% filter(!is.na(COMID))
nrow(data) #no rows removed

#Look at date range for all sites
summary(data$Date)

#Create year and month columns
data$year <- as.numeric(format(data$Date, "%Y"))
data$month<- as.numeric(format(data$Date, "%m"))

#Look at months during which samples were collected
data %>% group_by(month) %>% count()

#Omit samples collected in winter (December, January, February, March)
#(i.e., outside of typical range during which biotic condition is assessed)
nrow(data)
data<-data %>% filter(!month %in% c(1,2,3,12))
nrow(data)


#Break out the huc2 regions 
levels(factor(data$HUC4))
data$huc2 <-as.numeric(substr(as.character(data$HUC4), 1, nchar(as.character(data$HUC4))-2))

data$huc2
levels(factor(data$huc2))

huc2_names <- cbind(unique(data$huc2[order(data$huc2)]),
                    as.data.frame(c("New England", "Mid-Atlantic", "South Atlantic Gulf", 
                                    "Great Lakes", "Ohio", "Tennessee", "Upper Mississippi", 
                                    "Lower Mississippi", "souris Red Rainy", "Missouri", 
                                    "Arkandas-White-Red", "Texas-Gulf", "'Rio Grande", 
                                    "Upper Colorado","Lower Colorado", "Great Basin", "Pacific-Northwest", "California")))
colnames(huc2_names) <- c("huc2", "huc2name")

head(data)

data <- merge(data, huc2_names, by="huc2", all.x=FALSE)

nrow(data)


#Add unique numeric site ID based on unique location:

data2<-data %>% 
  group_by(Lat, Long) %>% 
  mutate(LatLong_ID = cur_group_id()) %>%
  ungroup()

View(data2)


#Count total number of samples per site/date - are there subsamples?

data3<-data2 %>% 
  add_count(LatLong_ID, Date, name="samples_per_date")

data3 %>% filter(samples_per_date>1) %>% count()/
nrow(data3)

#am seeing some subsample data
#~10% of all rows are multi samples per date (appears more applicable for river samples)

#Average BCG subsamples collected on same day by same agency
data3<-data3 %>% 
  group_by(LatLong_ID, Date, State) %>% 
  mutate(BCG_daily=mean(BCG_proxy)) %>% 
  ungroup() %>% 
  dplyr::select(-c(BCG_proxy)) %>%  #drop subsamples collected on the same day 
  distinct() %>% 
#now add counts per site/year
  add_count(LatLong_ID, year, name="samples_per_year") %>% 
#If some sites have many samples per year, that will skew the average HUC8 average towards the BCG values at those sites
#(Same issue will occur if some sites are sampled repeatedly over the years; but in next module will select most recent yearly sample for every site)
  #create average BCG score for every site/year combination
  #this also deals with the same site/year being included in multiple agency datasets by averaging them
  group_by(LatLong_ID, year) %>% 
  mutate(BCG_yearly=mean(BCG_daily)) %>% 
  ungroup() %>% 
  dplyr::select(-c(BCG_daily, Date, samples_per_date, SiteID, State, month)) %>% 
  distinct() %>% 
#now add counts per site
  add_count(LatLong_ID, name = "total_years_visited")


View(data3)

#Are there still duplicates, for samples in the same year from multiple agency datasets?

data3 %>% 
  add_count(LatLong_ID, year, name="agency_duplicates") %>% 
  filter(agency_duplicates>1)

#No duplicates
```

# 4 Write clean data to file

Clean dataset - Filtered to 1 BCG Obs per Site per Year

```{r}

#Write cleaned data to file: 
write_xlsx(data3, path =  paste(dir$output, "/bcg_data_cleaned_one_score_per_year.xlsx", sep=""))

```
