---
title: "2.0 BCG spatial coverage for survey"
author: "Christy Dolph"
date: "2025-04-01"
output: html_document
---

# Overview

This script creates spatial layers and maps of Biological Condition Gradient (BCG) data for use in the EPA Round 2 Project survey.

BCG data was assembled and BCG proxies were previously assigned by Christy Dolph using script 'BCG_proxy_conversion.Rmd'

BCG data was previously cleaned and processed by Virginia Callison and Christy Dolph using script '2.0_Initial_BCG_data_QC.Rmd'.

Authors and contributors:

VC = Virginia Callison

CLD = Christy Dolph

# Workspace Setup

```{r, setup, include =FALSE}
knitr::opts_chunk$set(echo = TRUE)

# Clean up workspace and load or install packages if necessary
rm(list=ls())
want <- c("data.table", "fixest", "tidyverse", "tidylog", "readxl", "lubridate", "rmarkdown", "writexl", "sf", "mapview", "xtable", "pals", "tmap", "maps", "gridExtra", "ffmpeg", "magick", "animation", "geosphere", "tigris")
need <- want[!(want %in% installed.packages()[,"Package"])]
if (length(need)) install.packages(need)
lapply(want, function(i) require(i, character.only=TRUE))
rm(want, need)
```

Set up working directories:

```{r, Create list of directories}
dir <- list()

#Main directory: 
dir$root <- getwd()

#Directory for loading data from EPA Project Round 1:
dir$data_r1 <- paste(dir$root, "/EPA_Round_1_data", sep = "")
#Directory for loading data from EPA Project Round 2:
dir$data_r2 <- paste(dir$root, "/EPA_Round_2_data", sep = "")
#Directory for spatial data: 
dir$spatial_dir<-paste(dir$root, "/Spatial_data", sep="")
#Directory with HUC shapefiles: 
dir$data_huc  <- paste(dir$root, "/Spatial_data/HUC_spatial_data", sep = "")
#Output directory for most data: 
dir$output<-paste(dir$root, "/BCG_output", sep="")

```

# Load pre-processed BCG data:

This BCG data has had the following pre-processing:

-   Assembled from state, regional, and federal agencies by VC and CLD

-   Biological index scores converted to BCG "proxy" by CLD (see script: "BCG_conversion_script.R")

-   BCG sample locations merged to WBD and NHD attributes by CLD (see script "BCG_spatial_analysis_script.R")

-   Additional pre-processing by VC and CLD (see script "0.1_qc_REVISED.R"):

    -   Filtered to 1 Observation per Site per Year

    -   Sites that were sampled multiple times per year were averaged to create an average BCG score per year

```{r}
#Read in pre-processed data: 

bcg_data_filtered<-read_xlsx(paste(dir$output,
"/bcg_data_filtered.xlsx", sep=""))

head(bcg_data_filtered)
nrow(bcg_data_filtered)
names(bcg_data_filtered)
#View(bcg_data_filtered)
```

## Additional Data Processing:

-   Select only most recent year for each site

-   Select data only post 2000

-   Format HUC IDs

```{r}

#Select most recent year sampled for each site 
Recent<-bcg_data_filtered %>%  
  group_by(LatLong_ID) %>%
  slice(which.max(year)) %>% #select most recent year for each site
  ungroup() 
nrow(Recent)  

#Look at date range for most recent samples
summary(Recent$year) #note this includes some old samples
#restrict to since 2000

Recent<-Recent %>% 
  filter(year>2000)

View(head(Recent))

#Check HUC8 IDs for leading zero 
levels(factor(Recent$HUC8)) 

nrow(Recent)
names(Recent)
write_xlsx(Recent, path =  paste(dir$output, "/bcg_most_recent_since2000.xlsx", sep=""))

```

Calculate average BCG by HUC8

```{r}
#calculate average BCG by HUC8 (since 2000)
names(Recent)

HUC8.avg<-Recent %>% 
  #mutate(HUC8=sprintf("%08.0f", (HUC8)) %>% #NOTE: only include if need to add leading zero for IDs to get 8 digit ID 
  group_by(HUC8) %>% 
  summarize(BCG_avg=mean(BCG_yearly_avg), n=n()) %>% mutate(BCG_avg_round=round(BCG_avg, digits=0)) #create round number for average BCG 

nrow(HUC8.avg)

#Look at HUC8 watersheds with >=5 sample points
View(HUC8.avg %>% filter(n>=5))
#956/1464 watersheds have >= 5 sample points

#recheck HUC8 IDs, leading zeros now included nrow(HUC8.avg)
levels(factor(HUC8.avg$HUC8)) 

#write to file 
write_xlsx(HUC8.avg, path = paste(dir$output, "/BCG_avg_by_HUC8.xlsx", sep=""))
```

## Create spatial data for BCG sample locations

```{r}
#Converting to sf Object 

names(Recent)
BCG_sf <- st_as_sf(Recent, coords = c("Long","Lat")) 
BCG_sf
names(BCG_sf)

#Set coordinate reference system to wgs84 
st_crs(BCG_sf) = 4326 

st_bbox(BCG_sf) #bounding box #can ID spatial extent of dataset

#Do a quick check of spatial data
plot(BCG_sf["BCG_yearly_avg"])

```

## Create spatial data for average BCG by HUC8

```{r}

#read in HUC8 boundaries 

HUC8_shp <- st_read(paste(dir$spatial_dir, "/HUC_spatial_data/HUC8_CONUS/HUC8_CONUS.shp", sep="")) 

head(HUC8_shp)

st_bbox(HUC8_shp) #check bounding box

st_crs(HUC8_shp) #reference system is WGS84

#Merge average BCG information to HUC8s 
HUC.BCG<-merge(HUC8_shp, HUC8.avg, by="HUC8", all.x=TRUE) 
head(HUC.BCG)
nrow(HUC.BCG)

#Simplify columns
HUC.BCG.simple<-HUC.BCG %>% dplyr::select(HUC8, AREASQKM, STATES, NAME, BCG_avg_round, n, geometry) %>% 
  rename(BCG_AVG=BCG_avg_round) #make cleaner column name


View(HUC.BCG.simple)


names(HUC.BCG)

#Simple plot of HUC8s 
plot(HUC.BCG["BCG_avg_round"])
plot(st_geometry(HUC.BCG))


#Write Avg BCG by HUC8 to shapefile:
st_write(HUC.BCG.simple, paste(dir$spatial_dir, "/BCG_avg_by_HUC8.shp", sep=""), append=FALSE)

```

### Fancier map of avg BCG by HUC8

```{r}
#map HUC8s (using tmap) 
names(HUC.BCG.simple)

#tmap_options(check.and.fix = TRUE) #set tmap options

tm_shape(HUC.BCG.simple)+
  tm_polygons(fill="BCG_AVG")


#tm_layout(legend.outside = TRUE)+ #several more options available to clean up legend tm_shape(subset(World, name=="United States"))+ tm_borders()
```

### Look at new data for Florida

```{r}
names(HUC.BCG)

US<-st_read(paste(dir$spatial_dir, "/united_states.shp", sep=""))
st_crs(US)
US_WGS84<-st_transform(US, crs=st_crs(HUC.BCG.simple))

head(US_WGS84)

names(HUC.BCG.simple)

tm_shape(US %>% filter(STATE_NAME %in% c("Florida", "Georgia", "Alabama")))+
  tm_borders()+
tm_shape(st_make_valid(HUC.BCG))+
  tm_polygons(fill="BCG_avg_round")+
  tm_text("n", size=6)

```

### Look at new data for Connecticut

```{r}
tm_shape(US %>% filter(STATE_NAME %in% c("Connecticut")))+
  tm_borders()+
tm_shape(st_make_valid(HUC.BCG))+
  tm_polygons(fill="BCG_avg_round")+
  tm_text("n", size=6)
```

## Characterizing HUC4s for targeting surveys

-   HUC4 list

-   Number/list of HUC8s in each HUC4

-   Number of biological samples in each HUC8

-   Assigned BCG level for each HUC8 (if assignment exists)

-   Population density in each HUC8

I suggest we proceed in stages. Christy, if you can assemble the environmental elements, perhaps Yongjie can then fold in the population density figures from env atlas.

Here are the study area definitions:

-   [NW]{.underline}: HUC4s 1707, 1709, 1710, 1712, 1801, 1802

-   [SW]{.underline}: HUC4s 1503, 1504, 1505, 1506, 1507, 1810

-   [MW]{.underline}: all HUC4s in state of Nebraska plus 1012, 1014

-   [SE]{.underline}: HUC4s 0306, 0307, 0308, 0311, 0312, 0313, 0315

-   [NE]{.underline}:  HUC4s in NY, ,VT, NH, MA (perhaps not sampling NYC and Boston)

```{r}

#Load in HUC4 data:
HUC4<-st_read(paste(dir$data_huc, "/HUC4_CONUS.shp", sep=""))
names(HUC4)

#check reference system
st_crs(HUC4)

plot(HUC4["huc4"])
```

```{r}
#List of priority watersheds

#Named list of priority watersheds from Dan P:
HUC4.list_1<-c("1707", "1709", "1710", "1712", "1801", "1802", "1503", "1504", "1505", "1506", "1507", "1810", "1012", "1014", "0306", "0307", "0308", "0311", "0312", "0313", "0315")
length(HUC4.list_1)

#View(HUC4)

#Plus additional HUC4s in NE, NY, ,VT, NH, MA (perhaps not sampling NYC and Boston)
#read in USA state boundaries using tigris package
library(tigris)
states<-states()
states_sf <- st_as_sf(states) #convert to sf object
st_bbox(states_sf) #bounding box
states_sf

st_crs(states_sf) #check reference system

names(states_sf)

#Select states of interest
Select_states_sf<-states_sf %>% filter(STUSPS %in% c("NE", "NY", "VT", "NH", "MA"))
Select_states_sf
plot(Select_states_sf["STUSPS"])

#Select HUC4s that intersect these states:
# First need to use st_make_valid on HUC4

HUC4_valid<-st_make_valid(HUC4)

HUC4_select_2<-st_intersection(HUC4_valid, Select_states_sf)

#Check HUC4s
plot(HUC4_select_2["STUSPS"])

#Make list of these HUC4s:

HUC4.list_2<-HUC4_select_2$huc4
HUC4.list_2

#Merge two HUC4 lists into one
HUC4.list_priority<-c(HUC4.list_1, HUC4.list_2)
HUC4.list_priority

#Select Priority Watersheds
HUC4_PRIORITY<-HUC4 %>% filter(huc4 %in% HUC4.list_priority)

View(HUC4_PRIORITY)

#Write priority HUC4s to shapefile:
st_write(HUC4_PRIORITY, paste(dir$spatial_dir, "/HUC4_PRIORITY.shp", sep=""), append=FALSE)

plot(HUC4_PRIORITY["huc4"])
```

## Characterize HUC4s

-   HUC4 list

-   Number/list of HUC8s in each HUC4

-   Number of biological samples in each HUC8

-   Assigned BCG level for each HUC8 (if assignment exists)

-   Population density in each HUC8

```{r}

#Merge HUC4 ID to HUC8s (if not there already)

#First 4 digits of HUC8 are HUC4 ID

#Make a HUC4 column
HUC.BCG.simple$huc4<-substr(HUC.BCG.simple$HUC8, 1, 4)
View(HUC.BCG)


#Select HUC8s from priority HUC4s:

HUC.BCG.PRIORITY<-HUC.BCG.simple %>% filter(huc4 %in% HUC4.list_priority)

View(HUC.BCG.PRIORITY)

```

### Plot HUC4s with detailed BCG info

Color palletes: <https://github.com/cols4all/cols4all-R>

```{r}

#make valid geometry for HUC8s:

HUC.BCG.PRIORITY_valid<-st_make_valid(HUC.BCG.PRIORITY)

levels(factor(HUC.BCG.PRIORITY_valid$BCG_AVG))

#Write Avg BCG by HUC8 to shapefile, w/HUC4 ID:
st_write(HUC.BCG.PRIORITY_valid, paste(dir$spatial_dir, "/HUC.BCG.PRIORITY_valid.shp", sep=""), append=FALSE)

View(HUC.BCG.PRIORITY_valid)

#Create a label column for "n"
HUC.BCG.PRIORITY_valid$n_label <- ifelse(is.na(HUC.BCG.PRIORITY_valid$n), "", HUC.BCG.PRIORITY_valid$n)

View(HUC.BCG.PRIORITY_valid)

#View map interactively: 
tmap_mode("view")

HUC4.Map<-
tm_shape(HUC.BCG.PRIORITY_valid)+
  tm_fill(fill="BCG_AVG", 
              fill.scale = tm_scale_categorical(values = "area7"))+
              tm_text("n_label", size=1)+
tm_shape(HUC.BCG.PRIORITY_valid)+
  tm_borders(lwd=0.5, col="gray")+
  tm_shape(HUC4_PRIORITY)+
  tm_borders(lwd=2, col="black")
HUC4.Map


#Add ecoregion


tm_save(HUC4.map, paste(dir$maps, "Priority_HUC4s.png"))
```

Create tables to summarize HUC4 characteristics

```{r}

#ID HUC4s where all HUC8s have >=5 sample sites 

##LEFT OFF 4.3.25

#Number of HUC8s by HUC4

HUC.BCG.PRIORITY_valid %>% 
  st_drop_geometry %>% 
  group_by(huc4) %>% 
  count()


```

## Finer resolution mapping: Minnesota

Load in data obtained from Urban Biomonitoring Project (w/Claire Bass)

Data source: MPCA

(only a few local agencies have IBI scores available, and these are not included here yet)

```{r}

setwd("G:/Shared drives/Finlay lab/Projects/Urban_biomonitoring/Biomonitoring_datasets/MPCA")

#read in info about sampling stations
station<-read_excel("./ClaireBass_DataRequest.xlsx", sheet="Station Information")

head(station)

invert.data<-read_excel("./ClaireBass_DataRequest.xlsx", sheet="Invertbrate_Data_IndexScores")

head(invert.data)
```

Link Lat/Long to IBI scores:

```{r}
names(station)
names(invert.data)

invert.sites<-merge(invert.data, station, by=c("FieldNum"))
head(invert.sites)
```

Convert MN IBI scores to BCG proxy (based on criteria from Round 1 of EPA project - see Appendix of EPA paper)

Just realized that you didn't use proxies for MN BCGs, because you had actual BCG scores! And we don't have these for new urban biomonitoring request.

```{r}

```
