---
title: "3.0 BCG spatial coverage for survey"
author: "Christy Dolph"
date: "2025-04-01"
output: html_document
---

# Overview

This script creates spatial layers and maps of Biological Condition Gradient (BCG) data for use in the EPA Round 2 Project survey.

BCG data was assembled and BCG proxies were previously assigned by Christy Dolph using script '1.0_BCG_proxy_conversion.Rmd'

BCG data was previously cleaned and processed by Virginia Callison and Christy Dolph using script '2.0_Initial_BCG_data_QC.Rmd'.

Authors and contributors:

VC = Virginia Callison

CLD = Christy Dolph

# Workspace Setup

```{r, setup, include =FALSE}
knitr::opts_chunk$set(echo = TRUE)

# Clean up workspace and load or install packages if necessary
rm(list=ls())
want <- c("data.table", "fixest", "tidyverse", "tidylog", "readxl", "lubridate", "rmarkdown", "writexl", "sf", "mapview", "xtable", "pals", "tmap", "maps", "gridExtra", "ffmpeg", "magick", "animation", "geosphere", "tigris")
need <- want[!(want %in% installed.packages()[,"Package"])]
if (length(need)) install.packages(need)
lapply(want, function(i) require(i, character.only=TRUE))
rm(want, need)
```

Set up working directories:

```{r, Create list of directories}
dir <- list()

#Main directory: 
dir$root <- getwd()

#Directory for loading data from EPA Project Round 1:
dir$data_r1 <- paste(dir$root, "/EPA_Round_1_data", sep = "")
#Directory for loading data from EPA Project Round 2:
dir$data_r2 <- paste(dir$root, "/EPA_Round_2_data", sep = "")
#Directory for spatial data: 
dir$spatial_dir<-paste(dir$root, "/Spatial_data", sep="")
#Directory with HUC shapefiles: 
dir$data_huc  <- paste(dir$root, "/Spatial_data/HUC_spatial_data", sep = "")
#Output directory for most data: 
dir$output<-paste(dir$root, "/BCG_output", sep="")

```

# Load pre-processed BCG data:

This BCG data has had the following pre-processing:

-   Assembled from state, regional, and federal agencies by VC and CLD

-   Biological index scores converted to BCG "proxy" by CLD (see script: "BCG_conversion_script.R")

-   BCG sample locations merged to WBD and NHD attributes by CLD (see script "BCG_spatial_analysis_script.R")

-   Additional pre-processing by VC and CLD (see script "0.1_qc_REVISED.R"):

    -   Filtered to 1 Observation per Site per Year

    -   Sites that were sampled multiple times per year were averaged to create an average BCG score per year

```{r}
#Read in pre-processed data: 

bcg_data_filtered<-read_xlsx(paste(dir$output,
"/bcg_data_filtered.xlsx", sep=""))

head(bcg_data_filtered)
nrow(bcg_data_filtered)
names(bcg_data_filtered)
#View(bcg_data_filtered)
```

## Additional Data Processing:

-   Select only most recent year for each site

-   Select data only post 2000

-   Format HUC IDs

```{r}

#Select most recent year sampled for each site 
Recent<-bcg_data_filtered %>%  
  group_by(LatLong_ID) %>%
  slice(which.max(year)) %>% #select most recent year for each site
  ungroup() 
nrow(Recent)  

#Look at date range for most recent samples
summary(Recent$year) #note this includes some old samples
#restrict to since 2000

Recent<-Recent %>% 
  filter(year>2000)

View(head(Recent))

#Check HUC8 IDs for leading zero 
levels(factor(Recent$HUC8)) 

nrow(Recent)
names(Recent)
write_xlsx(Recent, path =  paste(dir$output, "/bcg_most_recent_since2000.xlsx", sep=""))

```

Calculate average BCG by HUC8

```{r}
#calculate average BCG by HUC8 (since 2000)
names(Recent)

HUC8.avg<-Recent %>% 
  #mutate(HUC8=sprintf("%08.0f", (HUC8)) %>% #NOTE: only include if need to add leading zero for IDs to get 8 digit ID 
  group_by(HUC8) %>% 
  summarize(BCG_avg=mean(BCG_yearly_avg), n=n()) %>% mutate(BCG_avg_round=round(BCG_avg, digits=0)) #create round number for average BCG 

nrow(HUC8.avg)

#Look at HUC8 watersheds with >=5 sample points
View(HUC8.avg %>% filter(n>=5))
#956/1464 watersheds have >= 5 sample points

#recheck HUC8 IDs, leading zeros now included nrow(HUC8.avg)
levels(factor(HUC8.avg$HUC8)) 

#write to file 
write_xlsx(HUC8.avg, path = paste(dir$output, "/BCG_avg_by_HUC8.xlsx", sep=""))
```

# Create spatial data for BCG sample locations

```{r}
#Converting to sf Object 

names(Recent)
BCG_sf <- st_as_sf(Recent, coords = c("Long","Lat")) 
BCG_sf
names(BCG_sf)

#Set coordinate reference system to wgs84 
st_crs(BCG_sf) = 4326 

st_bbox(BCG_sf) #bounding box #can ID spatial extent of dataset

#Do a quick check of spatial data
plot(BCG_sf["BCG_yearly_avg"])

```

# Create spatial data for average BCG by HUC8

```{r}

#read in HUC8 boundaries 

HUC8_shp <- st_read(paste(dir$spatial_dir, "/HUC_spatial_data/HUC8_CONUS/HUC8_CONUS.shp", sep="")) 

head(HUC8_shp)

st_bbox(HUC8_shp) #check bounding box

st_crs(HUC8_shp) #reference system is WGS84

#Merge average BCG information to HUC8s 
HUC.BCG<-merge(HUC8_shp, HUC8.avg, by="HUC8", all.x=TRUE) 
head(HUC.BCG)
nrow(HUC.BCG)

#Simplify columns
HUC.BCG.simple<-HUC.BCG %>% dplyr::select(HUC8, AREASQKM, STATES, NAME, BCG_avg_round, n, geometry) %>% 
  rename(BCG_AVG=BCG_avg_round) #make cleaner column name


View(HUC.BCG.simple)


names(HUC.BCG)

#Simple plot of HUC8s 
plot(HUC.BCG["BCG_avg_round"])
plot(st_geometry(HUC.BCG))


#Write Avg BCG by HUC8 to shapefile:
st_write(HUC.BCG.simple, paste(dir$spatial_dir, "/BCG_avg_by_HUC8.shp", sep=""), append=FALSE)

```
